---
title: "Analysis of severe weather data - Impact on public health and economy"
author: "Jakob Schelbert"
date: "November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
This report deals with the effects of severe weather events and gives some implications for future countermeasures.
Using the *Storm Data* from the National Weather Service, we investigate which types of weather events have the biggest impact on the health of the population and economic consequences.
The data was collected during the period from 1950 till 2011.
For this we use the raw data and compute some scores to compare each type of event which leads to a ranking.
The analysis shows that *floods* have the highest economic impact, while *tornados* pose the most severe threat towards public health.
The insights could be used to devise some countermeasures to reduce the impacts of the events.


## Used packages

```{r packages, message=FALSE}
library(knitr)
library(dplyr)
library(lubridate)
library(data.table)
library(ggplot2)
```


## Data Processing
We download the data (for an explaination of the raw data see [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)) from the website (or use the code below to download it) and load it into *R*.
```{r data-loading, cache = TRUE}
url <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
datafile <- "repdata_data_StormData.csv.bz2"
if (!file.exists(datafile)) {
  download.file(url, datafile)
}
  
data_raw <- read.csv("repdata_data_StormData.csv.bz2")
```

The raw data for the property and crop damages are stored by using a variable for the numeric value and one which expresses the exponent (e.g. *K* for a multiplication with 1000).
In order to have a concise tidy data we add two additional variables which hold the correct value of the damages taking the exponent into account.

```{r data-processing, cache = TRUE}
data <- data.table(data_raw)
data <- data %>% mutate(property_damage = 0.0, crop_damage = 0.0)

exponents_list <- list(list("K", 1000), list("M", 1000000), list("B", 1000000000), 
                       list("k", 1000), list("m", 1000000), list("b", 1000000000), 
                       list("h", 100), list("H", 100), list("", 1), list("0", 1),
                       list("1", 10), list("2", 10^2), list("3", 10^3), list("4", 10^4),
                       list("5", 10^5), list("6", 10^6), list("7", 10^7), list("8", 10^8),
                       list("9", 10^9), list("?", 1), list("+", 1), list("-", 1)
                       )

for(i in exponents_list) {
    exponent_identifier <- i[[1]]
    exponent_multiplier <- i[[2]]
    data[data$PROPDMGEXP == exponent_identifier, "property_damage"] <-
        data[data$PROPDMGEXP == exponent_identifier, "PROPDMG"] * exponent_multiplier
    data[data$CROPDMGEXP == exponent_identifier, "crop_damage"] <-
        data[data$CROPDMGEXP == exponent_identifier, "CROPDMG"] * exponent_multiplier
}
```

For the economic impact we use the sum of the property damage and the crop damage which is stored in the variable *economy_damage*
```{r economic_damage}
data <- data %>% mutate(economy_damage = property_damage + crop_damage)
```

Furthermore, to be able to compare the different weather events to each other, we use an aggregated score for harmfulness towards public health.
This score is computed from the number of deaths and injuries that originate from each event and is stored in the variable *health_impact_score*.
For the sake of simplicity we rate a death ten times more severe as an injury.
The reader might want to adjust the score to her/his needs.
```{r health_score}
death_factor <- 22.0
injury_factor <- 1.0
data <- data %>% mutate(health_impact_score = death_factor * FATALITIES + injury_factor * INJURIES)
```

### Interdependency between injuries and fatalities
Our weights for the health score is based on the observation that fatalities and injuries are correlated which we quickly discuss in the following section.
We calculate a simple linear regression between fatalities and injuries to see that these are indeed correlated.
```{r correlation_fatalities_injuries}
correlation_fatalities_injuries <- lm(FATALITIES ~ INJURIES, data = data)
summary(correlation_fatalities_injuries)
```
Here we see that about 22 (1/0.0453207 = 22.06497) injuries correspond to one fatality, hence the weights we defined above.


## Results
The results are split into two subsections.
The first deals with the economic impact while the second sheds some light on the health impact of the events.

### Economic impact
To get an overview of the economic impact of each type of weather event we group the data and compute some basic statistics for each type.
More precisely, we report for each type of events the overall *number* of events, the *sum* of the economic impact for all events in the data set, the *mean*, the *standard deviation*, and the *mean absolute deviation*.

```{r economic_impact}
eco_summary <-
    data %>% group_by(EVTYPE) %>% summarize(
    number_of_events = n(),
    sum_economic_damage = sum(economy_damage),
    mean_economic_damage = mean(economy_damage),
    sd_economic_damage = sd(economy_damage),
    mad_economic_damage = mad(economy_damage)
    )
```
One could put the emphasis on the sum of all economic damage that was caused by all occurences of one event type.
In the following we present the ten event types with the highest sum of economic damage.
```{r}
kable(head(eco_summary %>% arrange(desc(sum_economic_damage)), 10))
```

In this case we see that **floods** have the highest sum of economic damages in the data set.
A graphical representation is depicted in the following figure.
```{r eco_plot}
top_eco_events <- head(eco_summary %>% arrange(desc(sum_economic_damage)), 10)
g <- ggplot(data = top_eco_events, aes(x=reorder(EVTYPE, sum_economic_damage), y=sum_economic_damage))
g + geom_bar(stat = "identity") + coord_flip() + labs(title = "Events with highest economic impact", y="Economic damage [US $]", x = "Event type")
```


### Health impact
Analogously as for the economic impact we create a summary table of the public health related variables.
Here we report for each event type the *number* of events, the *sum*, the *mean*, the *standard deviation*, and the *mean absolute deviation* for fatalities, injuries and our above defined health impact score.
```{r health_impact}
health_summary <-
    data %>% group_by(EVTYPE) %>% summarize(
    number_of_events = n(),
    sum_deaths = sum(FATALITIES),
    sum_injuries = sum(INJURIES),
    mean_deaths = mean(FATALITIES),
    mean_injuries = mean(INJURIES),
    sd_deaths = sd(FATALITIES),
    sd_injuries = sd(INJURIES),
    mad_deaths = mad(FATALITIES),
    mad_injuries = mad(INJURIES),
    sum_score = sum(health_impact_score),
    mean_score = mean(health_impact_score),
    sd_score = sd(health_impact_score),
    mad_score = mad(health_impact_score)
    )
```

Again, we can consider a ranking based on the sum of the health impact score which results in the following table.
```{r}
kable(head(
    health_summary %>% arrange(desc(sum_score)) %>%
    select(EVTYPE,
           number_of_events,
           sum_deaths,
           sum_injuries,
           sum_score,
           mean_score),
    10
    ))
```

Here we see that **tornados** have the highest sum of the health impact score which corresponds to the high number of fatalities and injuries.
A graphical representation is depicted in the following figure.
```{r health_plot}
top_health_events <- head(health_summary %>% arrange(desc(sum_score)), 10)
g <- ggplot(data = top_health_events, aes(x=reorder(EVTYPE, sum_score), y=sum_score))
g + geom_bar(stat = "identity") + coord_flip() + labs(title = "Events with highest health impact", y="Health score", x = "Event type")
```


## Discussion and outlook
While this analysis gives some insights on the types of weather events with the highest impact in both economy and public health, a more thourough investigation should be done in order to decide which countermeasure should be put in place.
A time plot of the impact could be one point to see whether events have already been addressed in the near past.
In addition, we suggest to look for countermeasures that address more than one type of weather event to efficiently invest money.


## Appendix
For the sake of completeness we report the computer setup that was used for gerenating this report.
### System setup and used package versions
```{r}
sessionInfo()
```

