---
title: "Course Project 2 - Analysis of Storm Data"
author: "Jakob Schelbert"
date: "10 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Review criteria
* Has either a (1) valid RPubs URL pointing to a data analysis document for this assignment been submitted; or (2) a * lete PDF file presenting the data analysis been uploaded?
* Is the document written in English?
* Does the analysis include description and justification for any data transformations?
* Does the document have a title that briefly summarizes the data analysis?
* Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?
* Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for * ysis?
* Is there a section titled "Results" where the main results are presented?
* Is there at least one figure in the document that contains a plot?
* Are there at most 3 figures in this document?
* Does the analysis start from the raw data file (i.e. the original .csv.bz2 file)?
* Does the analysis address the question of which types of events are most harmful to population health?
* Does the analysis address the question of which types of events have the greatest economic consequences?
* Do all the results of the analysis (i.e. figures, tables, numerical summaries) appear to be reproducible?
* Do the figure(s) have descriptive captions (i.e. there is a description near the figure of what is happening in the * re)?
* As far as you can determine, does it appear that the work submitted for this project is the work of the student who submitted it?



## Synopsis
This report deals with the effects of severe weather events and gives some implications for future countermeasures.
Using the *Storm Data* from the National Weather Service, we investigate which types of weather events have the biggest impact on the health of the population and economic consequences.

Zusammenfassung...
Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?

## Used packages

```{r packages, message=FALSE}
library(dplyr)
library(data.table)
library(ggplot2)
```


## Data Processing
We download the data from the website (or use the code below to download it) and load it into *R*.
```{r data-processing}
url <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
datafile <- "repdata_data_StormData.csv.bz2"
if (!file.exists(datafile)) {
  download.file(url, datafile)
}
  
# TODO: Check fread possibility to load the zipped file!
data_raw <- read.csv("repdata_data_StormData.csv.bz2")

# TODO: Convert the costs to one exponent (i.e. 25K = 25.000)
data <- data.table(data_raw)
data <- data %>% mutate(property_damage = 0.0, crop_damage = 0.0)

exponents_list <- list(list("K", 1000), list("M", 1000000), list("B", 1000000000), 
                       list("k", 1000), list("m", 1000000), list("b", 1000000000), list("",1))

for(i in exponents_list){
  exponent_identifier <- i[[1]]
  exponent_multiplier <- i[[2]]
  data[data$PROPDMGEXP==exponent_identifier,"property_damage"] <- data[data$PROPDMGEXP==exponent_identifier,"PROPDMG"] * exponent_multiplier
  data[data$CROPDMGEXP==exponent_identifier,"crop_damage"] <- data[data$CROPDMGEXP==exponent_identifier,"CROPDMG"] * exponent_multiplier
}


```

In order to be able to compare the different weather events to each other, we use an aggregated score for harmfulness towards public health.
This score is computed from the number of deaths and injuries that originate from each event.
```{r health_score}
death_factor <- 5.0
injury_factor <- 1.0
data <- data %>% mutate(health_impact_score = death_factor * FATALITIES + injury_factor * INJURIES)
```

For the economic impact we use the sum of the property damage and the crop damage which is stored in the variable *economy_damage*
```{r economic_damage}
data <- data %>% mutate(economy_damage = property_damage + crop_damage)
```

## Results

```{r economic_impact}
eco_summary <- data %>% group_by(EVTYPE) %>% summarize(number_of_events = n(), sum_economic_damage = sum(economy_damage), mean_economic_damage = mean(economy_damage), sd_economic_damage = sd(economy_damage), mad_economic_damage = mad(economy_damage)) %>% arrange(desc(mean_economic_damage))
```



```{r health_impact}
health_summary <- data %>% group_by(EVTYPE) %>% summarize(number_of_events = n(), sum_deaths = sum(FATALITIES), sum_injuries = sum(INJURIES), mean_deaths = mean(FATALITIES), mean_injuries = mean(INJURIES), sd_deaths = sd(FATALITIES), sd_injuries = sd(INJURIES), mad_deaths = mad(FATALITIES), mad_injuries = mad(INJURIES), sum_score = sum(health_impact_score), mean_score = mean(health_impact_score), sd_score = sd(health_impact_score), mad_score = mad(health_impact_score))
```

## Appendix

### System setup and used package versions
```{r}
sessionInfo()
```

